---
title: "Exploratory Data Analysis - MovieLens"
author: "Manuel Alejandro Matías Astorga"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(readr)
```

### 📥 Load Data

```{r load-data}
ratings <- read_csv("../data/ml-1m/ratings.csv")
movies <- read_csv("../data/ml-1m/movies.csv")
users <- read_csv("../data/ml-1m/users.csv")
```

### 🎯 Ratings Distribution

```{r ratings-distribution}
ggplot(ratings, aes(x = rating)) +
  geom_bar(fill = "steelblue") +
  labs(title = "Distribution of Ratings", x = "Rating", y = "Count") +
  theme_minimal()
```

### 🎬 Most Rated Movies

```{r top-rated-movies}
ratings %>%
  count(movieId, sort = TRUE) %>%
  top_n(10) %>%
  left_join(movies, by = "movieId") %>%
  ggplot(aes(x = reorder(title, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkorange") +
  coord_flip() +
  labs(title = "Top 10 Most Rated Movies", x = "Movie", y = "Number of Ratings") +
  theme_minimal()
```

### 👤 Most Active Users

```{r active-users}
ratings %>%
  count(userId, sort = TRUE) %>%
  top_n(10) %>%
  ggplot(aes(x = reorder(as.factor(userId), n), y = n)) +
  geom_bar(stat = "identity", fill = "forestgreen") +
  coord_flip() +
  labs(title = "Top 10 Most Active Users", x = "User ID", y = "Ratings Count") +
  theme_minimal()
```

### 📊 Rating Matrix Sparsity (Heatmap idea)

```{r sparsity-sample}
library(Matrix)

# sample 100 users × 100 movies
rating_sample <- ratings %>%
  sample_n(5000) %>%
  pivot_wider(names_from = movieId, values_from = rating) %>%
  select(-userId)

# convert to matrix and plot sparsity
image(as.matrix(!is.na(rating_sample)), col = c("white", "black"),
      main = "User-Movie Interaction Sparsity (Sample)")
```

## ✍️ Summary and Next Steps

This initial exploratory analysis confirms the dataset's sparsity and user-item imbalance.
Most users rate few movies, while a handful of movies dominate the rating count.

To enhance our recommendation strategy, we will:

- Enrich the movies dataset using TMDb API (overview, genres, cast, etc.)
- Store the extended information in MongoDB Atlas
- Use this information for a content-based and hybrid recommendation model
